---
author: "YOUR NAME HERE"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

# Gridded and tabular data retrieval
**15 pnts**

## Objectives
 In this lab you will:
 -Retrieve flow data using the dataretrieval package
 -Learn to retrieve and manage digital elevation models
 -Retrieve and import remotely sensed data for the watershed extent

### Setup:<br>
When importing packages and setting parameters, think about how the local working environment can vary among users and computers. A function that examines and installs packages before importing them enhances the script's ability to be easily used by others (a.k.a., enhances reproducibility). 

```{r include=FALSE}
# Above, include = FALSE means this chunk of script will not show when you convert this Markdown document to an HTML.

pkgTest <- function(x)
{
  if (x %in% rownames(installed.packages()) == FALSE) {
    install.packages(x, dependencies= TRUE)
  }
  library(x, character.only = TRUE)
}

# Make a vector of the packages you need
neededPackages <- c('tidyverse', 'dplyr', 'dataRetrieval', 'raster', 'sp', 'sf', 'leaflet', 'terra', 'ggplot2', 'plotly') 
# dataRetrieval - access data from NWIS
# sf - open and manage .shp files
# leaflet - one option for interactive visualization of spatial data
# raster and terra - .tif file management 

# For every package in the vector, apply your pkgTest function
for (package in neededPackages){pkgTest(package)}
```

In this lab, we will use the USGS dataRetrieval package to search for and filter hydrologic datasets directly from the USGS Water Data system. This package provides a powerful interface for discovering monitoring locations, identifying available parameters, and downloading time series data.

The package has recently undergone updates, and several older NWIS functions are being replaced with newer read_waterdata_*() functions. If you encounter warnings referencing deprecated NWIS functions, this simply means the function has a newer equivalent. The most up-to-date guidance can be found in the official USGS tutorial:[dataRetrieval tutorial](https://https://waterdata.usgs.gov/blog/dataretrieval/).

If you currently use scripts that utilize older API's, and want more on the transition see: (https://waterdata.usgs.gov/blog/api-waterservices-decom/)

In this example, we will use the new functions identify USGS monitoring locations in Colorado that:

--Represent stream sites

--Fall within a defined watershed size range

--Report discharge (streamflow) data

--Contain records during a specified time period

The objective is not just to download data, but to learn how to determine whether data exist for a given site before building an analysis workflow. This is a critical step in developing independent research projects. By building your workflow directly from publicly available data using scripted queries, you ensure that your analysis is transparent, reproducible, and easily updated in the future.

```{r}
# Start by defining and storing your search parameters as variables at the beginning of the script. This practice enables you and other users to easily modify the search parameters from a single location within the script.

state_name <- "Colorado"  
gage_no <- 'USGS-09034000'
```

**Q1 (2pnts). If you were creating this script for your work team's shared collaboration, or for yourself to use again a year from now, what might be the disadvantages of embedding search parameters directly into the function parameters?** 

ANSWER: 


```{r}
# Let's see what sites are available within the state. This will return something like 39000 sites without a filter, so we'll filter by drainage area to give us a managable df.
sites <- read_waterdata_monitoring_location(state_name = state_name) %>% # filter by drainage area
    filter(drainage_area < 300 & drainage_area > 250) 
```

**Q2 (1pnt). In a sentence, what does this dataframe (sites) tell you?**

ANSWER: 


If you look at the 'sites' dataframe that is now in the Data section of your Global Environment. It should contain ~44 observations. Try clicking on the drop down arrow by the dataframe name to see the structure, or head() or str() to get a quick glance at the df without opening the entire table. You can see that we get some site information here, but not enough metadata to tell us what kind of data is being monitored/collected at the site. Let's use the read_watershed_ts_meta() function to pull metadata, and then filter by parameter. 

How can we figure out the parameter codes for the the variables that we want to filter by? 
```{r}
param_codes <-read_waterdata_parameter_codes()
```

There are a lot of options available here, though this does not mean these are available at every site. I just wanted you to see the kinds of variables you could search for when developing/refining your question for the term project. Since we know we want discharge, lets narrow down the scope of codes. 

```{r}
read_waterdata_parameter_codes(
  parameter_name = "Discharge"
) %>%
  dplyr::select(parameter_code, parameter_name, unit_of_measure)
```

**(1pt) Below, use the parameter code to collect sites that report in ft3/s** 
This is typically how discharge is reported by federal agencies in the U.S., though you will see hydrologists convert to metric when modeling or disseminating.

```{r}
#Now filter for sites containing discharge data:
sites_discharge <- sites %>%
  semi_join(
    read_waterdata_ts_meta(
      monitoring_location_id = sites$monitoring_location_id
    ) %>%
      filter(parameter_code == "00060") %>%  
     distinct(monitoring_location_id),
   by = "monitoring_location_id"
  )
```

Note that while your research interests may not require streamflow data, dataRetrieval can serve as a valuable resource for a wide array of data types. 


For the remainder of this assignment, we'll focus on a single watershed in central Colorado near the Fraser Experimental Forest. Lets use the read_waterdata_ts_meta again to see what data is available for a particular site number. Note that we saved gage_no for the desired site above when we defined parameter values for the script.

```{r}
# Check data availability for the gage of interest
dataavailable <- read_waterdata_ts_meta(monitoring_location_id = gage_no)
```

**Q3 (2pnts). From dataavailable,** a. Identify one parameter at this site that is not discharge (00060). What does it measure, and in what units?

b. Based on the begin and end dates of that parameter’s record, would you consider this dataset suitable for long-term trend analysis? Briefly justify your answer.


ANSWER: 

### Flow data retrieval:
For this, we are going to use the legacy function (NWIS). You will get a warning, that is fine. The newest data is not yet available for this site with the new functions. This could be the case with sites of interest for your projects as well. In that case, the legacy functions will work through the end of the semester. 
```{r}
# Raw daily data:
rawQData <- readNWISdv(
  siteNumbers = "06730200",
  parameterCd = "00060",
  startDate = "2014-01-01",
  endDate = "2024-01-01"
)
```

Let's rename our columns. For this you will need to determine the units of the date 

```{r}
rawQData <- rawQData %>%
  rename ('ft3sec' = 'X_00060_00003') %>%
# the X_00060_00003_cd column provides information about the data quality. It is good to check this to anticipate or understand potential data anomalies. We have checked this and determined that we no longer need it for this exercise:
  dplyr::select(-c('X_00060_00003_cd', 'agency_cd'))

```

```{r}
# It is good to visualize your data upon import:

ggplot(rawQData, aes(x = Date, y = ft3sec)) +
  geom_point() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 month") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Q4 (2pnts). Based on the hydrograph plot, what type of precipitation events appear to influence the observed intra-annual variations in flow rates?** 

ANSWER: 




### Watershed boundary vector data: 

A boundary shapefile has been provided to save time and improve repeatability. Watershed boundaries for gauging stations can be found at [USGS StreamStats](https://streamstats.usgs.gov/ss/) by zooming into the water map and clicking 'Delineate' near your desired gage or point. Once your watershed is delineated, you can click on 'Download Basin' and select from a choice of file types. This can be an easy to to delineate river basins if all you require is a shapefile, though there are a lot of tools available for further hydrological analysis (e.g. Whitebox, SAGA)...more on that later. We will cover those in a future module. 

Let's import a pre-created shapefile: 

```{r}
import_path <- paste(getwd(), '/cabin_creek_shp/globalwatershed.shp', sep='')

# Read the shapefile
ws_boundary <- st_read(import_path)
```

```{r}
# Check the coordinate reference system (CRS) of the shapefile
print(st_crs(ws_boundary))
```

Let's transform this to a projected CRS (flat surface) rather than a geographic (spherical surface) CRS. A projected CRS will depict the data on a flat surface which is required for distance or area measurements. A more detailed explanation can be [found at the University of Colorado's Earth Lab](https://www.earthdatascience.org/courses/earth-analytics/spatial-data-r/intro-to-coordinate-reference-systems/) if needed.

```{r}
# Use the EPSG code for UTM Zone 12. Choose a UTM or other code appropriate for the area of interest.
new_crs <- st_crs(32613)  # Use the appropriate EPSG code for your CRS
ws_boundary <- st_transform(ws_boundary, crs = new_crs)

# We will export the transformed shapefile for use in the MODIS gui later
st_write(ws_boundary, "cabin_creek_shp_projected/cabin_creek_projected.shp", append=FALSE)
```

#### Visualization - watershed boundary 
i.e., sanity check

```{r}
# Let's look at the shapefile with a basic leaflet map. Leaflet can be a nice way to see your data with on a basemap for geographic context.

# Transform the shapefile to long-lat for visualization
ws_boundary_longlat <- st_transform(ws_boundary, crs = st_crs("EPSG:4326"))

# Create the Leaflet map
map <- leaflet() %>%
  setView(lng = -105.7, lat = 40.0, zoom = 13) %>%
  addProviderTiles("Esri.WorldImagery") %>%
  addPolygons(data = ws_boundary_longlat, fillOpacity = 0.7, layerId = "my_layer")

# Print or display the map
map
```

### Digital Elevation Models (DEM)
A DEM has been provided for consistency among users.

However, if you are unfamiliar with the process of retrieving a DEM here are three methods that our team commonly uses:<br>

1. USDA's [Natural Resources Conservation Science (NRCS) website](https://datagateway.nrcs.usda.gov/GDGOrder.aspx). Here you can order (free) DEMs by state, county or bounding box. It takes just minutes for them to be emailed to you. <br><br>

2. If you have a USGS account (fast and free) you can also find DEM data at https://earthexplorer.usgs.gov. See [this video](https://youtube.com/watch?v=NQg0g9ObhXE) if interested. This is the method we used to provide this DEM and we will use this resource again to retrieve Landsat data.<br><br>

3. GIS or hydrological software like QGIS and SAGA offer tools for downloading DEMs directly into your project for analysis. This can be optimal when working over large geographic extents to streamline the data acquisition process. Both QGIS and SAGA are free and easy to learn if you have experience with ESRI products.<br><br>

Let's check out our DEM:

```{r}
# Import the pre-downloaded DEMs. Note that we need two files to obtain a complete DEM for our watershed.
dem1 <- raster(file.path('cabin_creek_dems_raw/n39_w106_1arc_v3.tif'))
dem2 <- raster(file.path('cabin_creek_dems_raw/n40_w106_1arc_v3.tif'))

# Check that the DEMs align. If retrieving them from different sources, you can also check that the projections are the same with proj4string() before mosaicing
origin(dem2) <- origin(dem1)

# Mosaic the rasters
dem <- mosaic(dem1, dem2, fun = mean) # fun indicates the treatment of overlapping cells

# Project the raster to match the shapefile's projection
dem_projected <- projectRaster(dem, crs = crs(ws_boundary))

# Check
plot(dem_projected)
plot(ws_boundary, add = TRUE)
```

If the watershed and raster don't overlap as you expect, be sure the projection and extent is the same for all data files plotted. 

```{r}
# Clip the dem to the watershed. This will change all elevation values outside of the watershed to 'NA'
dem_clipped <- mask(dem_projected, ws_boundary) 

# Remove rows and columns with only 'NA' values
dem_trimmed <- trim(dem_clipped)

# Save the trimmed raster to a new file or use it for further analysis
writeRaster(dem_trimmed, "trimmed_dem.tif", format = "GTiff", overwrite=TRUE)

plot(dem_trimmed)
```

### Other remotely sensed data:

Now that we have site data spatial data, let's cover a couple of techniques or resources you can use to access satellite data for your site. In the event that you only need a vegetative map layer from a specific point in time, it is not too difficult to download actual Landsat imagery from usgs.gov. In the case of Cabin Creek let's look at a single summer image from 2020.<br><br>

Navigate to https://ers.cr.usgs.gov/, and if you don’t already have an account, you can create one quickly or sign in if you already have one. Once signed in, go to https://earthexplorer.usgs.gov you should see a surface map with tabs on the left. 
Start with the ‘Search Criteria’ tab. For this assignment the shape file we have is too large to upload, so it is easiest to scroll down and choose ‘polygon’ or ‘circle’ to define your area. Under the circle tab you can use Center Latitude 39.9861   Center Longitude -105.719 and a radius of about 3.5 kilometers to capture our study area. 

Let's say we want an NDVI map layer from summer of 2020. In 'date range', select dates for the summer of 2020. In this high elevation area of Colorado, the greenest times of the year without snow cover are likely June and July with senescence beginning sometime in late July. First search from June 1 to July 15 2020. Then click on ‘Data Sets’ either at the bottom or top of the left-hand side menus. From the data sets we will select Landsat Collection 2 level 2, Landsat 8 - 9 OLI TIRS C2L2. 

We can skip ‘Additional Criteria’ for now and go to ‘Results’. You should receive a set of about 11 images. For each image you can check out the metadata. That is the 'pen and notepad' icon in the middle of image specific menu. Here you can look at data like cloud cover. For these images, the values you see are a cloud coverage percent. Cloud coverage measurements and the codes to filter for certain cloud cover thresholds may change across missions. If you are retrieving Landsat data for time series using code, it will be important to explore the different quality control parameters and what they represent. You don't need to download anything for this assignment, we've provided it for you. However to gain familiarity with this process, click on the 'download' icon for July 7, 2020 and click 'select files' on the Reflectance Bands option. 


**Q5 (1pnt): What 2 bands should we download to calculate NDVI?**

Answer:



Check out other earthexplorer resources while you are at the site. You will likely find resources here for your MS projects, even if it is just for making maps of your study area for presentations or paper.

##### Landsat 
If you want to learn about Landsat missions or utilize Landsat data, USGS offers all of the reference information you need. The sensors on these missions can be a fantastic resource if characterizing land or water surface over large temporal scales. Landsat data provides (30m) spatial resolution so can capture spatial heterogeneity, even for smaller study areas. If calculating vegetation indices over different time periods, it is important to know which mission you are retrieving data from and check the band assignments for that mission. For example, on the Landsat 6 instrument, Band 4 is near-infrared, whereas Band 4 on Landsat 9's instrument is red. 

##### Moderate Resolution Imaging Spectroradiometer (MODIS):
is a sensor aboard NASA's Terra and Aqua satellites which orbit Earth and collect data on a daily basis. MODIS data provides moderate spatial resolution (250m to 1km) and high temporal resolution. Therefore, MODIS data can be an optimal way to monitor or characterize rapid changes to Earth's surface like land cover changes resulting from wildfire. Data can be accessed and downloaded directly through [NASA's Earthdata tool](https://www.earthdata.nasa.gov/).


Let's import our raster data. 

```{r}
# Import the two bands needed for NDVI calculation
red_band <- raster("landsat/LC08_L2SP_034032_20200702_20200913_02_T1_SR_B4.TIF")
nir_band <- raster("landsat/LC08_L2SP_034032_20200702_20200913_02_T1_SR_B5.TIF")

# Note that the file name tells us about the data.

# Calculate NDVI
ndvi <- (nir_band - red_band) / (nir_band + red_band)

# Project the NDVI layer so that the coordinate reference system matches that of the watershed boundary shapefile.
ndvi_projected <- projectRaster(ndvi, crs = crs(ws_boundary))

# Crop NDVI raster to the extent of ws_boundary. 
ndvi_crop <- crop(ndvi_projected, ws_boundary)

# Plot NDVI
plot(ndvi_crop, col = rev(rainbow(255)))

# Plot ws_boundary as an outline
plot(ws_boundary, add = TRUE, border = "black", lwd = 2, col = "transparent")
```


**Q6 (1pnt)  What does 'SR' in the .tif names tell us about the data?**

ANSWER: 


**Q7 (2pnts)  What is the difference between 'mask' and 'crop' when filtering large files to fit the area of interest?**


ANSWER: 


#### Earth surface changes over time

If considering changes over time, comparing individual rasters from different time periods directly may not always provide meaningful insights because the absolute values of vegetation indices can vary due to factors such as differences in solar angle, atmospheric conditions, and land cover changes. Instead, it's often more informative to analyze trends in vegetation over time.

##### LandTrendr
One fun way to do this is with [Oregon State University's LandTendr](https://emapr.github.io/LT-GEE/landtrendr.html). With a Google account and access to Earth Engine, you can easily interact with the UI Applications in section 8. You can jump to the [pixel timeseries here](https://emaprlab.users.earthengine.app/view/lt-gee-pixel-time-series).
This will allow you to view changes in vegetation indices over a specified time period within a single Landsat pixel. 

##### Google Earth Engine
Another effective way to access evaluate large spatial datasets is through Google Earth Engine (GEE). GEE is a cloud-based platform developed by Google for large scale environmental data analysis. It provides a [massive archive of satellite imagery and other geospatial datasets](https://developers.google.com/earth-engine/datasets/), as well as computational tools to view and analyze them. GEE uses Google's cloud to provide processing of data so you don't need powerful computing ability to run a potentially computationally expensive spatial analysis. 

###### GEE Access

If spatial data is something you may need for your dissertation or MS project, we recommend exploring this resource.

**Sign Up:** Go to the Google Earth Engine website (https://earthengine.google.com/) and sign in with your Google account. If you don't have a Google account, you can sign up for access by filling out a request form.

**Request Access:** If you're signing up for the first time, you'll need to request access to Google Earth Engine. Provide information about your affiliation, research interests, and how you plan to use Earth Engine (learn and access data!).

**Approval:** After submitting your request, you'll need to wait for approval from the Google Earth Engine team. Approval times may vary, but you should receive an email notification once your request has been approved.

**Access Earth Engine:** Once approved, you can access Google Earth Engine through the Code Editor interface in your web browser at code.earthengine.google.com to start exploring the available datasets, running analyses, and visualizing your results.

We have provided a [basic script for retrieving and viewing NDVI data](
https://code.earthengine.google.com/332cc9055781b21d8506620f0a4cb582). This is a code snapshot, so you can manually copy the code from the snapshot and paste it into you own Google Earth Engine Code editor. 
You are welcome to save this to your own GEE files and try changing the area of interest, vegetation indices, time periods or exploring other data. 

Colorado State University also has a [great tutorial here](https://ecology.colostate.edu/google-earth-engine/) that will start you from scratch. 

Note that JavaScript is the default language for the code editor interface in your web browser. If you are familiar with Python, GEE provides a Python API that can allow you to use Python syntax. If neither of these languages are familiar, don't give up yet! You can still use this tool with your R coding background and trial and error. We recommend finding other scripts that suit your need (there are a lot of shared resources online) and adapting them to your area of interest. Once you understand coding concepts, like those in R, other languages can be picked up easily. You will find that you are learning JavaScript before you realize it. 

**Q8 (4 pts) Using the monitoring data tools (e.g., USGS dataRetrieval) and/or spatial data platforms (e.g., Google Earth Engine, LandTrendr), describe one relationship or pattern you are curious about exploring (1–2 sentences). This does not need to be your final project question, just a realistic relationship that could be investigated within a few weeks using available data.**

In your response, include:

What type of data would be needed (e.g., discharge, NDVI, land cover, climate, water quality, burn severity, etc.)

How can we help you move forward?
 For example: Are you unsure what platform hosts the data?
              Do you need help determining whether data exist?
              Are you uncertain how to organize or analyze the data?
              Is there a technical skill you need to build?

Be honest and specific. The goal is to start shaping a feasible term project and identify where support will be most useful.



